---
title: "Data Science Specialization"
subtitle: "Reproducible Research: Peer Assessment 1"
author: "aliki m."
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

###Problem statement

The problem consists of data recorded on a personal activity device (such as Nike Fuelband, Fitbit, Jawbone Up, etc). The data is a .csv file, made up of 17.568 observations collected between the months of October and November, of the following variables:

1. number of steps taken in a 5-minute interval;
2. date on which the measurement was taken;
3. identifier for the 5-minute interval in which the measurement was taken.

This is an Rmarkdown document processed by Knitr in R version 3.2.1 on OSx Yosemite 10.10.5.

###PART I: loading and pre-processing data

The following code chunk generates an R script to be sourced separately, if desired.
```{r setup, include=TRUE}
library("knitr")
knit_hooks$set(purl = hook_purl)
```

The file can be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip). Download the .zip file to *activity.zip* and unzip the .csv file containing the aforementioned data.

```{r download}
if(!file.exists("activity.zip")){
fileUrl<-"http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile='activity.zip', method='curl')
dateDownloaded<-date()
}
unzip('activity.zip', files="activity.csv", exdir=".",unzip="internal")
```

We then load the data into a data frame called *df* using the **read.csv** command, as follows:

```{r loaddata}
df<-read.csv("activity.csv",header=TRUE,sep=",")
```

The following outputs the class of each variable in the data frame:

```{r df_class}
 sapply(df,class)
```

We observe that the class of *steps*, *date*, and *interval* is output as 'integer', 'factor', and 'integer', respectively. In the following R code chunk, we convert the class of the date variable into 'date' with format YYYY-MM-DD.

```{r classchange}
df$date<-as.Date(df$date, format="%Y-%m-%d")
```

###PART II: What is the mean total number of steps taken per day?

1. Make a histogram of the total number of steps taken each day

We introduce a new variable, *dfcomplete* in which missing values are omitted.

```{r noNA}
dfcomplete<-na.omit(df)
```
Here, we output the number of days of collected data we have as well as the number of days without any missing values (given by NA):
```{r numberdays}
length(unique(df$date))
length(unique(dfcomplete$date))
```
We have `r length(unique(df$date))` days of collected data and `r length(unique(dfcomplete$date))` days without NA values.

Next, we collect the total number of steps corresponding to each of the `r length(unique(dfcomplete$date))` days found in the *dfcomplete* data frame and output to a data frame called *outdf*. The latter has `r length(unique(dfcomplete$date))` observations of two variables: date and total number of steps in each day.

```{r aggregatesteps}
outdf <- aggregate(steps ~ date, dfcomplete,FUN=sum)
str(outdf)
```

In this assignment we use the **ggplot2** graphics package to generate all plots. Next, we graph one-dimensional data corresponding to the number of steps taken each data using a histogram. The bins are all set to an equal size of 1000 steps. 

```{r histogram1}
library("ggplot2")
qplot(outdf$steps, geom="histogram",xlab="number of steps",alpha=I(0.8),col=I("blue"),fill=I("orange"),binwidth=1000,main="Histogram of the total number of steps per day")
```

2. Calculate the mean and the median total number of steps taken each day.

We now calculate some descriptive statistics of the data set in *outdf* and in particular the mean and the median total number of steps that were taken each day.
```{r stats}
mean(outdf$steps)
median(outdf$steps)
```
The mean is 10766, rounded to the nearest integer and the median is `r median(outdf$steps)`.

###PART III: What is the average daily pattern?

1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r dailyactivity}
 daily.activity<- aggregate(steps ~ interval,dfcomplete,FUN=mean)
 ggplot(daily.activity, aes(x =daily.activity$interval, y = daily.activity$steps))+geom_line(color="blue")+xlab("5-minute interval")+ylab("average number of steps taken")+ggtitle("Averaged number of steps by interval")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r maxinterv}
 maxtab<-daily.activity[which.max(daily.activity$steps),]
 as.numeric(as.character(maxtab[[1]]))
```
 
The maximum number of steps is contained in interval `r as.numeric(as.character(maxtab[[1]]))`. 

###PART IV: Imputing missing values
 
 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
 
```{r missingvalues}
 sum(is.na(df))
```
 The number of missing values is `r sum(is.na(df))`.
 
 2. Devise a strategy for filling in all of the missing values in the dataset.
 
 
```{r impute}
library("dplyr")
df$weekdays<-weekdays(df$date) 
dfcomplete$weekdays<-weekdays(dfcomplete$date) 
meanperday<-as.data.frame(with(dfcomplete, tapply(steps, list(interval, weekdays), mean)))
dfNA<-df[which(is.na(df)),]
for (i in 1:length(unique(dfNA$weekdays))){
    dfNA$steps<-replace(dfNA$steps, dfNA$weekdays==unique(dfNA$weekdays)[i],meanperday[unique(dfNA$weekdays)][[i]])
}
newdf<-rbind(dfcomplete, dfNA)%>%arrange(-desc(date))
```

```{r histogram2+stats}
newdf2 <- aggregate(steps ~ date, newdf,FUN=sum)
qplot(newdf2$steps, geom="histogram",xlab="number of steps",alpha=I(0.8),col=I("blue"),fill=I("orange"),binwidth=1000,main="Histogram of the total number of steps per day (missing values imputated)")
mean(newdf2$steps)
median(newdf2$steps)
``` 

##Part V: Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r weekend|weekday}
weekend<-c("Saturday","Sunday")
newdf$dayfactor<-as.factor(ifelse(newdf$weekdays %in% weekend,"weekend","weekday"))
```

2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r dayfactoractivity}
daily.activity2<-aggregate(steps~interval+dayfactor,newdf,FUN=mean)
ggplot(daily.activity2, aes(x =interval, y = steps))+geom_line(color="blue")+xlab("5-minute interval")+ylab("average number of steps taken")+ggtitle("Averaged number of steps by interval")+facet_grid(dayfactor~.)
```